D3064R0: Bringing the Laws of Physics to Bear on the OOTA Problem

### 2024-01-10 ootavacuous.tex
January 10, 2024 update to paper title
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Add appendix and links for non-trivial semantic dependencies
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\begin{document}
\title{D3064R0: Is OOTA Execution Vacuous for Real-World C++ Implementations?}

@@

Worse yet, suppose that both \co{y} and \co{z} are zero.
Then there is no semantic dependency from either \co{y} or \co{z} to
\co{x} unless both \co{y} and \co{z} change.
Alternatively, when both \co{y} and \co{z} are zero, there is no semantic
dependency from either \co{y} or \co{z} to \co{x}, but rather a semantic
dependency from the combination of \co{y} \emph{and} \co{z} to \co{x}.
This means that precise definitions of sdep cannot always rely on
single-variable perturbations.

@@

Note also that many users would consider it to be an extremely unfriendly
act for the C++ implemenation to combine loads or stores to adjacent
objects where each object \co{is_lock_free()}, but the combination is not.

@@

This section has shown that an implementation that correctly handles
the memory model, observable behavior, and the as-if rule cannot form
OOTA cycles from volatile atomics when running on real systems in this
universe.
It has also presented a substantial list of optimizations that avoid
creation of OOTA cycles from non-volatile atomics.

@@

This paper has refined the definition of ``OOTA cycle'' from
\co{acyclic(sdep; rf)} to \co{acyclic(sdep; rfe)}, which enables the
application of temporal reasoning to OOTA analysis of the C++ memory
model.
This reasoning shows that correct real-world C++ implementations relying
on single-threaded analyses are inherently incapable of producing OOTA
cycles in UB-free executions of programs using volatile atomics.

@@

@@

The manual analysis uses the informal rule that a semantic dependency
exists between a set of relaxed atomic loads and a store on any execution
where different values loaded would change the address stored to, the
value stored, or the whether or not that store would be executed in
the first place, with the proviso that stores of the same value to the
same address having the same ordering properties are fungible.

@@

>>>>

\begin{document}
\title{D3064R0: Bringing the Laws of Physics to Bear on the OOTA Problem}

@@

Worse yet, suppose that both \co{y} and \co{z} are zero.
Then there is no semantic dependency from either \co{y} or \co{z} to
\co{x} unless both \co{y} and \co{z} change.
Alternatively, when both \co{y} and \co{z} are zero, there is no semantic
dependency from either \co{y} or \co{z} to \co{x}, but rather a semantic
dependency from the combination of \co{y} \emph{and} \co{z} to \co{x}.
This means that precise definitions of sdep cannot always rely on
single-variable perturbations.
See Appendix~\ref{app:Non-Trivial Semantic Dependencies}
for examples and additional discussion.

@@

Note also that many users would consider it to be an extremely unfriendly
act for a C++ implementation to combine loads or stores to adjacent
objects where each object \co{is_lock_free()}, but the combination is not.

@@

This section has shown that an implementation that correctly handles
the memory model, observable behavior, and the as-if rule cannot form
OOTA cycles from volatile atomics when running UB-free programs on real
systems in this universe.
It has also presented a substantial list of optimizations that can be
applied to non-volatile atomics without the risk of creating OOTA cycles.

@@

This paper has refined the definition of ``OOTA cycle'' from
\co{acyclic(sdep; rf)} to \co{acyclic(sdep; rfe)}, which enables the
application the laws of physics (and thus of temporal reasoning) to OOTA
analysis of the C++ memory model.
This reasoning shows that correct real-world C++ implementations relying
on single-threaded analyses are inherently incapable of producing OOTA
cycles in UB-free executions of programs using volatile atomics.

@@

This paper also uncovers some shortcomings in typical definitions of
``semantic dependency'', pointing out examples where a single semantic
dependency extends from multiple loads to a store, but not from any one
of those loads to that same store.
We therefore propose that the definition of semantic dependency be
extended accordingly.

@@

The manual analysis uses the informal rule that a semantic dependency
exists between a set of relaxed atomic loads\footnote{
	Please see
	Appendix~\ref{app:Non-Trivial Semantic Dependencies}
	for examples showing why this is a set of loads rather than a
	single load.}
and a store on any execution where different values loaded would change
the address stored to, the value stored, or the whether or not that store
would be executed in the first place, with the proviso that stores of
the same value to the same address having the same ordering properties
are fungible.

@@

\subsubsection{Non-Trivial Semantic Dependencies}
\label{app:Non-Trivial Semantic Dependencies}

Most of the examples in this paper involve simple semantic dependencies
connecting a single load to a single store.
One exception appears in
Appendix~\ref{app:Non-LB Litmus Tests},
but this section will present a more difficult example.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-mult-0.litmus @@
\caption{OOTA Multiplication Example, Initial Value Zero}
\label{lst:OOTA Multiplication Example, Initial Value Zero}
\end{listing}

Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
shows an example that builds on the multiplication-by-zero
discussion in
Section~\ref{sec:Semantic Dependencies Are Per Execution}.
Line~11 of \co{P0()} stores to \co{z} the product of the values loaded from
\co{x} and \co{y} on lines~9 and~10, respectively.
Similarly, lines~16 and~17 store to \co{x} and \co{y}, respectively,
the value that line~15 loads from \co{z}.
Because all three variables are initialized to zero, changes to the value
loaded from \co{x} cannot affect the value stored by line~11 to \co{z}
unless there is also a change loaded from \co{y} and vice versa.
However, if both \co{x} and \co{y} change, the value stored to \co{z}
will change.

Therefore, as noted in
Section~\ref{sec:Semantic Dependencies Are Per Execution},
there is no semantic dependency from the load from \co{x} to the
store to \co{z}, nor is there a semantic dependency from the load
from \co{y} to the store to \co{z}.
However, there \emph{is} a semantic dependency from the \emph{combination}
of the loads from \co{x} and \co{y} to the store to \co{z}.

This situation poses a challenge to current work on semantic dependencies,
which typically restricts such dependencies to a single load and store.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-mult-1.litmus @@
\caption{OOTA Multiplication Example, Initial Value 1}
\label{lst:OOTA Multiplication Example, Initial Value of One}
\end{listing}

In contrast, consider
Listing~\ref{lst:OOTA Multiplication Example, Initial Value of One},
which differs from
Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
only in the initial values (1 instead of 0) and the \co{exists}
clause (0 instead of 1).

Because the values loaded from \co{x} and \co{y} on lines~9 and~10
are non-zero, changes to either \co{x} or \co{y} will affect the
value stored to \co{z} by line~11.
There is therefore a separate semantic dependency from the load from
\co{x} to the store to \co{z} and from the load from \co{y} to the store
to \co{z}, which matches typical definitions of semantic dependency.
Yet
Listing~\ref{lst:OOTA Multiplication Example, Initial Value of One}
and
Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
can be thought of as being merely two different executions of the same
program.

One might ask why semantic dependencies can be non-trivial when hardware
address, control, and data dependencies are much more straightforward.
Answers include:
(1)~Respecting dependencies even when they reduce to a constant,
(2)~Less-aggressive optimizations, and
(3)~Lack of undefined behavior.

This situation further underscores the per-execution nature of semantic
dependencies as well as the need for an improved definition of ``semantic
dependency''.
We propose expanding the traditional definition to include dependencies
extending from groups of loads to a single store.
